---
title: "Selecting oat parents for the New York environment via BrAPI + genomic prediction"
author: "jeanlucj"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5)
```

## Overview

This pipeline pulls oat data from **T3/Oat** over **BrAPI** and ends in a ranked list of accessions to use as parents for crossing, targeted at the **New York growing environment**. The steps are:

1.  Find the **training** trials (within a radius of a New York center point, or an explicit `TRAINING_TRIALS` list) plus any `TEST_TRIALS`, tagged by role.
2.  Pull phenotypes and field design, and split them into training vs test sets.
3.  Identify the genotyping projects / protocols covering the training **and** test accessions, build a combined relationship matrix, and **subset it to the prediction targets** (training ∪ predictable test; `TEST_ACCESSIONS` included).
4.  **Stage 1** — per-trial adjusted means (BLUEs) from the training trials only.
5.  **Stage 2** — multi-environment genomic prediction (BGLR) → GEBVs for the target accessions, including those never phenotyped.
6.  Rank candidates with a selection index and flag redundant (closely related) picks.

All parameters live in `code/config.R`, which reads the long selection lists (training/test trials, `TEST_ACCESSIONS`, traits) from `data/config/*.txt`. Each step caches its pull/result under `data/`; set `refresh = TRUE` (or delete the cache) to re-run a step.

T3/Oat requires a login: copy `.Renviron.example` to `.Renviron` and fill in `T3_USERNAME` / `T3_PASSWORD` before building this report. `connect_t3()` reads that file itself, so it works in the non-interactive `wflow_build()` subprocess.

```{r source-code}
library(tidyverse)
purrr::walk(
  c("config.R", "01_connect.R", "02_find_trials.R", "03_get_phenotypes.R",
    "04_find_genotyping.R", "05_stage1_blues.R", "06_stage2_genomic_prediction.R",
    "07_select.R"),
  ~ source(here::here("code", .x))
)
conn <- connect_t3()
```

## 1. Phenotyping trials in the New York region

```{r trials}
trials <- find_ny_trials(conn)
cat(sum(trials$role == "training"), "training and", sum(trials$role == "test"),
    "test trial(s).\n")
trials |>
  count(role, locationName, countryCode, distance_km = round(distance_km)) |>
  arrange(role, distance_km) |>
  knitr::kable(caption = "Trials by role and location (nearest first)")
```

```{r trials-map}
trials |>
  count(longitude, latitude, name = "n_trials") |>
  ggplot(aes(longitude, latitude)) +
  geom_point(aes(size = n_trials), alpha = 0.6) +
  annotate("point", x = CENTER_LON, y = CENTER_LAT, colour = "red", shape = 4, size = 4) +
  labs(title = "Trial locations (red x = NY center)", size = "n trials") +
  theme_minimal()
```

## 2. Phenotypes and field design

```{r pheno}
pheno <- get_phenotypes(conn, trials$studyDbId)
sets  <- split_by_role(pheno, trials)   # train_pheno, train_acc, test_acc
cat("Observations:", nrow(pheno$pheno),
    "| training accessions:", nrow(sets$train_acc),
    "| test-trial accessions:", nrow(sets$test_acc), "\n")
pheno$pheno |>
  group_by(trait) |>
  summarise(n_obs = n(), n_studies = n_distinct(studyDbId),
            mean = mean(value), sd = sd(value), .groups = "drop") |>
  knitr::kable(caption = "Trait summary")
```

## 3. Genotyping coverage and combined relationship matrix

With `GENO_PROTOCOL_ID = NULL` every protocol that genotypes our training **and** test accessions is downloaded and its genomic relationship matrix is EM-combined into one (large GBS panels are thinned to \~`TARGET_DENSITY` markers; undownloadable ones are skipped). A precomputed pedigree relationship matrix (`PEDIGREE_DIR`) is stitched in when it overlaps our accessions. The combined matrix is then **subset to the prediction targets** (training + predictable test); training accessions with neither genotypes nor pedigree are injected with a prior-only diagonal.

```{r geno}
geno <- find_and_get_genotypes(conn, sets$train_acc, sets$test_acc, TEST_ACCESSIONS)
knitr::kable(head(geno$projects, 10),
             caption = "Genotyping PROJECTS ranked by accession coverage")
knitr::kable(head(geno$protocols, 10),
             caption = "Genotyping PROTOCOLS ranked by accession coverage")
cat("Protocols combined:", paste(geno$protocol_ids, collapse = ", "), "\n")
cat("Prediction GRM:", nrow(geno$G), "x", ncol(geno$G), "accessions\n")
```

## 4. Stage 1 — per-trial BLUEs

Stage 1 uses **training-trial phenotypes only** (`sets$train_pheno`); test-trial phenotypes are held out of model fitting.

```{r stage1}
blues <- stage1_blues(sets$train_pheno)
cat("BLUEs:", nrow(blues), "across", n_distinct(blues$studyDbId), "studies\n")
head(blues) |> knitr::kable(caption = "Example Stage-1 BLUEs")
```

## 5. Stage 2 — genomic prediction (BGLR)

```{r stage2}
gebv <- stage2_gblup(blues, geno)
gebv |>
  group_by(trait) |>
  summarise(n_candidates = n(),
            n_phenotyped = sum(phenotyped),
            n_predicted_only = sum(!phenotyped), .groups = "drop") |>
  knitr::kable(caption = "GEBV candidates per trait")
```

```{r cv}
cv <- map_dfr(unique(gebv$trait), ~ cv_accuracy(blues, geno, .x),
              .progress = pb_wrap("Cross-validation: traits"))
knitr::kable(cv, caption = "5-fold cross-validated prediction accuracy")
```

## 6. Breeders' output

The selection index is a raw, weighted sum of per-trait GEBVs (`TRAIT_WEIGHTS`); GEBVs are **not** standardized, since they are already appropriately shrunk given their uncertainty. The full table (all predicted accessions, plus a per-trait weight/accuracy block) is written to `output/breeders_output.csv`.

```{r select}
sel <- select_parents(gebv, cv = cv)
head(sel$parents, 20) |>
  knitr::kable(caption = "Top candidates by selection index (GEBVs un-standardized)")
sel$traits |>
  knitr::kable(caption = "Per-trait index weight and cross-validated accuracy")
cat("Full breeders' output written to", output_path("breeders_output.csv"), "\n")
```
