---
title: "Selecting oat parents for the New York environment via BrAPI + genomic prediction"
author: "jeanlucj"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5)
```

## Overview

This pipeline pulls oat data from **T3/Oat** over **BrAPI** and ends in a ranked
list of accessions to use as parents for crossing, targeted at the **New York
growing environment**. The steps are:

1. Find phenotyping trials at locations within a radius of a New York center point.
2. Pull phenotypes and field design for those trials.
3. Identify the genotyping projects / protocols that cover the trial accessions,
   and download marker data for the best-covering protocol.
4. **Stage 1** — per-trial adjusted means (BLUEs) accounting for field design.
5. **Stage 2** — multi-environment genomic prediction (BGLR) → GEBVs for every
   genotyped candidate, including those never phenotyped in the NY region.
6. Rank candidates with a selection index and flag redundant (closely related) picks.

All parameters live in `code/config.R`. Each step caches its pull/result under
`data/`; set `refresh = TRUE` (or delete the cache) to re-run a step.

```{r source-code}
library(tidyverse)
purrr::walk(
  c("config.R", "01_connect.R", "02_find_trials.R", "03_get_phenotypes.R",
    "04_find_genotyping.R", "05_stage1_blues.R", "06_stage2_genomic_prediction.R",
    "07_select.R"),
  ~ source(here::here("code", .x))
)
conn <- connect_t3()
```

## 1. Phenotyping trials in the New York region

```{r trials}
trials <- find_ny_trials(conn)
cat(nrow(trials), "phenotyping trials within", RADIUS_KM, "km of",
    sprintf("(%.2f, %.2f)\n", CENTER_LAT, CENTER_LON))
trials |>
  count(locationName, countryCode, distance_km = round(distance_km)) |>
  arrange(distance_km) |>
  knitr::kable(caption = "Trials by location (nearest first)")
```

```{r trials-map}
ggplot(trials, aes(longitude, latitude)) +
  geom_point(aes(size = after_stat(count)), stat = "count", alpha = 0.6) +
  annotate("point", x = CENTER_LON, y = CENTER_LAT, colour = "red", shape = 4, size = 4) +
  labs(title = "Trial locations (red x = NY center)", size = "n trials") +
  theme_minimal()
```

## 2. Phenotypes and field design

```{r pheno}
pheno <- get_phenotypes(conn, trials$studyDbId)
cat("Observations:", nrow(pheno$pheno),
    "| traits:", n_distinct(pheno$pheno$trait),
    "| accessions:", nrow(pheno$accessions), "\n")
pheno$pheno |>
  group_by(trait) |>
  summarise(n_obs = n(), n_studies = n_distinct(studyDbId),
            mean = mean(value), sd = sd(value), .groups = "drop") |>
  knitr::kable(caption = "Trait summary")
```

## 3. Genotyping coverage and combined relationship matrix

With `GENO_PROTOCOL_ID = NULL` every protocol that genotypes our accessions is
downloaded and its genomic relationship matrix is EM-combined into one (large GBS
panels are thinned to ~`TARGET_DENSITY` markers; undownloadable ones are skipped).
A precomputed pedigree relationship matrix (`PEDIGREE_DIR`) is stitched in when it
overlaps our accessions.

```{r geno}
geno <- find_and_get_genotypes(conn, pheno$accessions)
knitr::kable(head(geno$projects, 10),
             caption = "Genotyping PROJECTS ranked by accession coverage")
knitr::kable(head(geno$protocols, 10),
             caption = "Genotyping PROTOCOLS ranked by accession coverage")
cat("Protocols combined:", paste(geno$protocol_ids, collapse = ", "), "\n")
cat("Combined GRM:", nrow(geno$G), "x", ncol(geno$G), "accessions\n")
```

## 4. Stage 1 — per-trial BLUEs

```{r stage1}
blues <- stage1_blues(pheno$pheno)
cat("BLUEs:", nrow(blues), "across", n_distinct(blues$studyDbId), "studies\n")
head(blues) |> knitr::kable(caption = "Example Stage-1 BLUEs")
```

## 5. Stage 2 — genomic prediction (BGLR)

```{r stage2}
gebv <- stage2_gblup(blues, geno)
gebv |>
  group_by(trait) |>
  summarise(n_candidates = n(),
            n_phenotyped = sum(phenotyped),
            n_predicted_only = sum(!phenotyped), .groups = "drop") |>
  knitr::kable(caption = "GEBV candidates per trait")
```

```{r cv}
cv <- map_dfr(unique(gebv$trait), ~ cv_accuracy(blues, geno, .x))
knitr::kable(cv, caption = "5-fold cross-validated prediction accuracy")
```

## 6. Selected parents

```{r select}
selected <- select_parents(gebv, geno = geno, n_select = 20)
selected |>
  filter(selected) |>
  knitr::kable(caption = "Top candidates for crossing (selection index)")
cat("Full ranking written to", output_path("selected_parents.csv"), "\n")
```
